On Optimality of Finite Element Integration

### 2015-12-09 TOMS_Paper/assembly.tex
ADd stuff to sharing elimination algo..

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} details the steps for performing sharing elimination on a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest. It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration expressions. Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions, to ensure the uniqueness of operands. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.

We perform a second depth-first visit of the expression tree to track all occurrences of multilinear operands (i.e., any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression in normal form).  We create disjoint partitions of multilinear operands such that no spatial sharing is present among them. For each partition, we have two alternatives: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding the expression to expose spatial sharing and enable a subsequent factorization. These two alternatives basically correspond to points 1) and 2) in Section~\ref{sec:se-rln}.

Given a partition $P$, to decide between (i) and (ii) we adopt the two following criteria:
\begin{enumerate}
\item We compare the cardinality of $P$ to that of the set of multilinear symbols (e.g., test functions) in $P$. If the former is lower, we use (i); if the latter is lower, we use (ii); if they are identical, we check the next criterion, which attempts to optimize for the outer loops.
\item We know how many times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression. Let $n_{o}$ and $n_{s}$ denote these quantities. Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols. Let $c_{o}$ be the cost of $o$ (i.e., the number of operations). Assume, for this example, that $P$ depends on $[L_i, L_k]$ (which we know, from Figure~\ref{code:loopnest}, to have extent $M$ and $O$, respectively). We observe that the increase in operation count at the level of the outer loops would be: $M (O \sum_{o \in X} c_{o} + \sum_{o \in X} (n_{o} - 1))$ if (i) were applied; $M \sum_{s \in Y} (n_{s} - 1)$ if (ii) were applied. We then take (i) or (ii) depending on which quantity is minimum.
\end{enumerate}

On a last depth-first visit of the expression tree, we apply point (2) above to all of the $L_i$-dependent expressions.

%TODO what partition do we take first ? we need a partial ordering here
\end{Algo}

>>>>

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest. It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration. Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
In the following, we assume $|.|$ to be a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop. The cost, in number of operations, of a sub-expression $o$ is represented by $c_o$.

The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions. This ensures the uniqueness of operands. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.

We perform a second depth-first visit of the expression tree to build a partitioning of multilinear operands; that is, any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression in normal form. We partition the multilinear operands into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$. The multilinear operands in a partition $P_i$ share the same set of symbols (e.g., all operands in which the first order derivatives of test functions appear), while there is no sharing across different partitions. We have to establish:
\begin{enumerate}
\item the order in which sharing elimination will be applied to the partitions in $\mathbb{P}$ (this has a relevant impact on the outcome of the transformation process).
\item a scheduling strategy for each $P \in \mathbb{P}$.
\end{enumerate}

We start with point 1). We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{L_j}$ and $\mathbb{P}^{L_k}$, which logically separates partitions based on the loop they depend on. We then order the partitions lexicographically, with the position of partition $P^{L_l} \in \mathbb{P}^{L_l}$ given by ${<}|\mathbb{P}^{L_l}|,\ |L_l| \sum_{o \in P^{L_l}} c_o{>}$.

%TODO need intuition here!

We now tackle point 2). For each partition $P \in \mathbb{P}$, we have to decide between two options, as already explained in Section~\ref{sec:se-rln}. We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion. For this choice, we adopt the following criteria:
\begin{enumerate}
\item Let $S_P$ be set of multilinear symbols in $P$ (e.g., test functions). If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops
\item We know how many times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression. Let $n_{o}$ and $n_{s}$ denote these quantities. Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols. Assume, for this example, that $P$ depends on $[L_i, L_k]$ (which we know, from Figure~\ref{code:loopnest}, to have extent $M$ and $O$, respectively). We observe that the increase in operation count at the level of the outer loops would be: $M (O \sum_{o \in X} c_{o} + \sum_{o \in X} (n_{o} - 1))$ if (i) were applied; $M \sum_{s \in Y} (n_{s} - 1)$ if (ii) were applied. We then take (i) or (ii) depending on which quantity is minimum.
\end{enumerate}

%TODO need intuition here!

It remains to schedule $L_i$-dependent expressions; that is, operations that, logically, do not belong to the multilinear loop nest. We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above. In doing this, we make the assumption that an $L_i$-dependent expression is the result of a structured tensor operation. If this were not the case, the factorization is chosen greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
\end{Algo}
